---
title: "County Crosswalks Between Any Years"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{County Crosswalks Between Any Years}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE)
```

Most geographic change in the census system happens on decennial schedules --
but counties don't wait for the decade. Since 2000, counties and county
equivalents have been created, merged, split, renamed, and transferred between
one another a dozen-plus times. A county panel from, say, 2014 to 2023 that
ignores these events will silently misalign Alaska, South Dakota, Virginia,
and all of Connecticut.

The `crosswalk` package curates these events from the Census Bureau's
[Substantial Changes to Counties](https://www.census.gov/programs-surveys/geography/technical-documentation/county-changes.html)
documentation and ships them as internal package data, so county -> county
crosswalks work between **any** pair of years from 2000 onward -- offline,
with no API key.

```{r setup}
library(crosswalk)
library(dplyr)
```

## What's covered

| Event | Type | First vintage with new codes |
|---|---|---|
| Clifton Forge, VA gives up independent-city status | Merge | 2002 |
| Broomfield County, CO created | Split (from parts of four counties) | 2002 |
| Southeast Alaska reorganizations (Skagway, Hoonah-Angoon, Wrangell, Petersburg, Prince of Wales) | Splits/merges | 2008-2009 |
| York County / Newport News, VA territory exchange | Part transfer | 2008 |
| Bedford City, VA rejoins Bedford County | Merge | 2014 |
| Petersburg Borough, AK created | Split/merge | 2014 |
| Shannon County, SD renamed Oglala Lakota County (46113 -> 46102) | FIPS change | 2015 |
| Wade Hampton, AK renamed Kusilvak (02270 -> 02158) | FIPS change | 2015 |
| Valdez-Cordova, AK split into Chugach and Copper River | Split | 2020 |
| Connecticut counties replaced by nine planning regions | Reorganization | 2022 |

A subtlety worth internalizing: the years above are Census Bureau *product
vintages*, not effective dates. The Valdez-Cordova split took legal effect on
January 2, 2019, but 2019-vintage ACS products still use the old county; the
new codes first appear in 2020-vintage products. The package keys every event
to the first product vintage that uses the post-change codes (verified
empirically against ACS county inventories), because that's what determines
which codes appear in the data you're crosswalking.

## Basic usage

Request any forward year pair. The result composes all events in the interval
on top of an identity mapping over the full county universe -- unchanged
counties appear with an allocation factor of 1, so the crosswalk is never
sparse:

```{r county-basic}
county_14_23 <- get_crosswalk(
  source_geography = "county",
  target_geography = "county",
  source_year = 2014,
  target_year = 2023,
  silent = TRUE)

crosswalk_14_23 <- county_14_23$crosswalks$step_1

nrow(crosswalk_14_23)

## the counties that changed between 2014 and 2023
crosswalk_14_23 |>
  filter(source_geoid != target_geoid)
```

Renames (like 46113 -> 46102) carry an allocation factor of 1 with an
`identity` weighting factor. Splits allocate the source county across its
successors with population-based factors:

```{r split-example}
county_19_21 <- get_crosswalk(
  source_geography = "county",
  target_geography = "county",
  source_year = 2019,
  target_year = 2021,
  silent = TRUE)

## Valdez-Cordova (02261) split into Chugach (02063) and Copper River (02066)
county_19_21$crosswalks$step_1 |>
  filter(source_geoid == "02261")
```

Applying these crosswalks works exactly as described in
`vignette("crosswalk")`: pass the result to `crosswalk_data()` with your
county-level data.

## Forward-only, with one exception

County temporal crosswalks require `source_year < target_year`:

```{r backward-error, error = TRUE}
get_crosswalk(
  source_geography = "county",
  target_geography = "county",
  source_year = 2021,
  target_year = 2019)
```

The asymmetry is inherent: composing a *merge* forward is just addition, but
reversing it would require disaggregating the merged county's values back into
its components, which a crosswalk row cannot honestly express without
additional data. Rather than return allocation factors that look precise but
aren't, the package asks you to standardize *forward* to the latest year in
your panel.

## Sub-county geographies

County changes ripple into the GEOIDs of everything nested inside a county:
when Shannon County became Oglala Lakota, every tract, block group, and block
in `46113...` was relabeled to `46102...`. The package serves these relabels
for same-decade year pairs from 2010 onward:

```{r tract-relabel}
tract_14_19 <- get_crosswalk(
  source_geography = "tract",
  target_geography = "tract",
  source_year = 2014,
  target_year = 2019,
  silent = TRUE)

## identity rows for all unchanged tracts, relabels for the affected ones
tract_14_19$crosswalks$step_1 |>
  filter(source_geoid != target_geoid) |>
  head()
```

Because these sub-county changes are exact 1:1 relabels, they are also the
one case where *backward* temporal crosswalks are offered (e.g., tract
2019 -> 2014) -- reversing a relabel is lossless in a way reversing a merge
is not.

Year pairs that span a decennial census (e.g., tract 2014 -> 2023) can't be
pure relabels, because tract boundaries themselves changed in 2020. For those,
`get_crosswalk()` automatically plans a chain that routes through NHGIS
cross-decade crosswalks (and so requires an `IPUMS_API_KEY`):

```{r cross-decade, eval = FALSE}
## planned automatically as: 2014 tracts -[relabel]-> 2010-decade vintage
## -[NHGIS]-> 2020-decade vintage -[relabel]-> 2023 tracts
tract_14_23 <- get_crosswalk(
  source_geography = "tract",
  target_geography = "tract",
  source_year = 2014,
  target_year = 2023)
```

## Connecticut, briefly

Connecticut's 2022 replacement of its eight counties with nine
planning-region county equivalents is the most sweeping recent change --
every Connecticut county GEOID changed, and the old and new units don't nest.
Direct 2020 <-> 2022 requests are served by crosswalks from the CT Data
Collaborative (county-level requests fetch county GEOIDs via `tidycensus` and
so need a `CENSUS_API_KEY`); county requests spanning 2022 (like the
2014 -> 2023 example above) incorporate the same change through the curated
event registry, with population-based allocation factors where old counties
split across planning regions.

## Where the data comes from

The event registry, GEOID mappings, and year-by-year county universes are
built by `data-raw/build_county_events_sysdata.R` from Census Bureau
documentation and ACS county inventories, and ship as internal package data
(`R/sysdata.rda`). See
[`data-raw/README.md`](https://github.com/UI-Research/crosswalk/blob/main/data-raw/README.md)
for provenance details.
