---
title: "Getting Started with crosswalk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with crosswalk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
## chunks that pull data from the Census API only evaluate when a key is set
has_census_key <- Sys.getenv("CENSUS_API_KEY") != ""

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE)
```

## Overview

Census geographies change. Tracts are redrawn every decade, counties are
occasionally renamed, merged, or split, and many analyses need to move data
between geographies (say, from ZIP Code Tabulation Areas to PUMAs) that don't
nest within one another. A *crosswalk* is a table that relates source
geographies to target geographies, with allocation factors describing what
share of each source unit belongs to each target unit.

The `crosswalk` package provides a consistent interface to crosswalks from
four sources, plus tooling to apply them to your data:

| Source | What it covers | API key |
|---|---|---|
| **GeoCorr** (Missouri Census Data Center) | Same-year crosswalks between geographies (GeoCorr 2022 for 2020s geography; GeoCorr 2018 for 2010s geography) | None |
| **IPUMS NHGIS** | Cross-decade crosswalks (e.g., 2010 tracts to 2020 tracts) | `IPUMS_API_KEY` |
| **CT Data Collaborative** | Connecticut's 2020-2022 change from counties to planning regions | `CENSUS_API_KEY` (county requests only) |
| **County events** (curated, ships with the package) | County redefinitions outside decennial censuses, supporting county crosswalks between any years from 2000 on | None |

Two functions do most of the work:

- `get_crosswalk()` fetches crosswalk(s), automatically chaining multiple
  sources when both geography and year change
- `crosswalk_data()` applies crosswalk(s) to your data

This vignette walks through the core workflow. The examples that query the
Census Bureau's API require a `CENSUS_API_KEY`; if you're missing outputs
below, that's why.

```{r setup}
library(crosswalk)
library(dplyr)
```

## Fetching your first crosswalk

Suppose we have county-level data and need it at the PUMA (Public Use
Microdata Area) level. Counties don't nest within PUMAs, so we need a
crosswalk. With no years specified, `get_crosswalk()` uses GeoCorr 2022
(2020-Census geography):

```{r fetch-first}
## save fetched crosswalks to a directory so repeat requests read from disk;
## see the Caching section at the end of this vignette
cache_dir <- file.path(tempdir(), "crosswalk-cache")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

county_to_puma <- get_crosswalk(
  source_geography = "county",
  target_geography = "puma",
  weight = "population",
  cache = cache_dir)
```

`get_crosswalk()` always returns a list with three elements:

| Element | Description |
|---|---|
| `crosswalks` | A named list of crosswalk tibbles (`step_1`, `step_2`, ...) |
| `plan` | The plan describing each step: its source, geographies, and years |
| `message` | A human-readable description of the crosswalk chain |

```{r inspect-structure}
county_to_puma$message

county_to_puma$crosswalks$step_1 |>
  head()
```

Every crosswalk uses the same standardized columns, whatever its source:

| Column | Description |
|---|---|
| `source_geoid`, `target_geoid` | Identifiers for the source and target geographies |
| `allocation_factor_source_to_target` | The share of the source unit's `weight` allocated to the target unit |
| `weighting_factor` | The attribute used to compute allocation factors (e.g., `population`, `housing_all`, `land`, `identity`) |

A county entirely inside one PUMA has a single row with an allocation factor
of 1; a county split across PUMAs has one row per PUMA, with allocation
factors that sum to 1. Crosswalks always contain exactly one row per
source-target pair.

The `weight` argument ("population", "housing", or "land") determines *how*
split units are apportioned -- population-weighted, housing-unit-weighted, or
area-weighted. Match it to what you're crosswalking: population-based weights
for people-denominated variables, housing for housing-unit variables. See
`vignette("how-interpolation-works")` for details.

### Where did this crosswalk come from?

Each crosswalk tibble carries a `crosswalk_metadata` attribute documenting its
provenance -- the source, the parameters used, download URLs, citation
guidance, and more (see `?get_crosswalk` for the full structure):

```{r metadata}
metadata <- attr(county_to_puma$crosswalks$step_1, "crosswalk_metadata")

names(metadata)

metadata$data_source_full_name
```

## Applying a crosswalk to data

`crosswalk_data()` applies a crosswalk to your data. Let's fetch county-level
poverty counts from the American Community Survey and translate them to PUMAs:

```{r acs-pull, eval = has_census_key}
county_poverty <- tidycensus::get_acs(
    year = 2023,
    geography = "county",
    output = "wide",
    variables = c(below_poverty = "B17001_002"),
    progress_bar = FALSE) |>
  select(
    source_geoid = GEOID,
    count_below_poverty = below_povertyE)

head(county_poverty)
```

Note the column naming: `crosswalk_data()` detects what to do with each column
from its prefix.

| Prefix | Treatment |
|---|---|
| `count_` | Multiplied by the allocation factor, then summed (counts: population, households, loans, ...) |
| `mean_`, `median_`, `percent_`, `ratio_` | Weighted mean, using the allocation factor as the weight (rates, averages) |

If you'd rather not rename your columns, pass them explicitly via
`count_columns` and `non_count_columns`. The column holding your GEOIDs is
named by `geoid_column` (default `"source_geoid"`).

```{r apply, eval = has_census_key, message = TRUE}
puma_poverty <- crosswalk_data(
  data = county_poverty,
  crosswalk = county_to_puma)

head(puma_poverty)
```

The result has one row per *target* geography: the target identifier is named
`geoid` (plus `geography_name` when the crosswalk provides it).

## Join quality: did everything match?

Data rows whose GEOIDs don't appear in the crosswalk can't be allocated
anywhere, so they're dropped -- silently losing data is the main hazard of
crosswalking, which is why `crosswalk_data()` reports on it. The join-quality
messages print as it runs (suppress them with `show_join_quality = FALSE`),
and the same statistics are always attached to the result as the
`join_quality` attribute:

```{r join-quality, eval = has_census_key}
join_quality <- attr(puma_poverty, "join_quality")

## what share of data GEOIDs failed to match the crosswalk?
join_quality$pct_data_unmatched

## which GEOIDs were they? (none, in this case)
head(join_quality$data_geoids_unmatched)
```

See `?crosswalk_data` for the attribute's full structure, and
`vignette("how-interpolation-works")` for how to diagnose imperfect joins.

## Multi-step crosswalks: changing geography *and* year

No single source provides, say, 2014 counties to 2022 PUMAs. When both the
geography and the year change, `get_crosswalk()` plans a chain: first change
the year (holding geography constant), then change the geography at the target
year.

```{r multi-step}
county14_to_puma22 <- get_crosswalk(
  source_geography = "county",
  target_geography = "puma",
  source_year = 2014,
  target_year = 2022,
  weight = "population",
  cache = cache_dir)

county14_to_puma22$message
```

You don't need to do anything special to apply a chain --
`crosswalk_data()` applies each step in sequence:

```{r multi-step-apply, eval = has_census_key}
county_poverty_2014 <- tidycensus::get_acs(
    year = 2014,
    geography = "county",
    output = "wide",
    variables = c(below_poverty = "B17001_002"),
    progress_bar = FALSE) |>
  select(
    source_geoid = GEOID,
    count_below_poverty = below_povertyE)

puma22_poverty <- crosswalk_data(
  data = county_poverty_2014,
  crosswalk = county14_to_puma22)

head(puma22_poverty)
```

The attributes on the returned tibble describe the final step; pass
`return_intermediate = TRUE` to get each step's result (with its own
attributes) alongside the final one.

### Cross-decade crosswalks (NHGIS)

Crosswalks that span a decennial census -- 2010 tracts to 2020 tracts, for
example -- come from IPUMS NHGIS and require a free API key. Request one at
<https://account.ipums.org/api_keys> and store it in your `.Renviron` as
`IPUMS_API_KEY` (e.g., via `usethis::edit_r_environ()`). Then:

```{r nhgis, eval = FALSE}
tract_2010_to_zcta_2020 <- get_crosswalk(
  source_geography = "tract",
  target_geography = "zcta",
  source_year = 2010,
  target_year = 2020,
  weight = "population")

# Step 1: 2010 tracts -> 2020 tracts (NHGIS)
# Step 2: 2020 tracts -> 2020 ZCTAs (GeoCorr)
```

For a complete worked example -- standardizing six years of mortgage data to
2020 tract definitions -- see `vignette("standardizing-longitudinal-data")`.

## County crosswalks between any pair of years

Counties are occasionally redefined *outside* decennial censuses: renames
(Shannon County, SD becoming Oglala Lakota County in 2015), merges (Bedford,
VA in 2013), splits (Valdez-Cordova, AK in the 2020 vintage), and Connecticut's
2022 switch to planning regions. The package curates these events from Census
Bureau documentation, so county -> county crosswalks work between **any** pair
of years from 2000 onward (forward in time only), with no API key and no
network access:

```{r county-events}
county_changes <- get_crosswalk(
  source_geography = "county",
  target_geography = "county",
  source_year = 2014,
  target_year = 2023,
  silent = TRUE)

## most counties are unchanged (identity rows); here are some that changed
county_changes$crosswalks$step_1 |>
  filter(source_geoid != target_geoid) |>
  head()
```

See `vignette("county-crosswalks")` for what's covered and how these
crosswalks are constructed.

## Nested geographies: when you don't need a crosswalk

If your source geography nests exactly within your target (tracts within
counties, block groups within tracts), there's nothing to crosswalk -- every
source unit belongs to exactly one target. `get_crosswalk()` warns and returns
an empty crosswalk; just aggregate your data directly:

```{r nested, warning = TRUE}
nested <- get_crosswalk(
  source_geography = "tract",
  target_geography = "county")
```

## What combinations are supported?

`get_available_crosswalks()` lists every supported combination and which
source serves it:

```{r available}
available <- get_available_crosswalks()

available |>
  count(crosswalk_source)

## e.g., everything that translates data out of 2010 tracts
available |>
  filter(source_geography == "tract", source_year == 2010) |>
  head()
```

One nuance: GeoCorr combinations are listed under their reference years (2022
and 2018), but requests resolve by *decade* -- asking for 2021 tracts -> 2021
ZCTAs is served by the tract -> zcta 2022 row. See
`?get_available_crosswalks`.

## Caching

We passed `cache = cache_dir` to each `get_crosswalk()` call above, so each
component crosswalk was saved as a CSV the first time it was fetched:

```{r caching}
list.files(cache_dir)

## a repeat call with the same parameters reads from disk instead of
## re-downloading
refetched <- get_crosswalk(
  source_geography = "county",
  target_geography = "puma",
  weight = "population",
  cache = cache_dir,
  silent = TRUE)

attr(refetched$crosswalks$step_1, "crosswalk_metadata")$read_from_cache
```

In a real project, use a persistent directory (e.g.,
`here::here("crosswalks-cache")`) rather than `tempdir()`. Cached files never
expire; delete a file to force a re-download.

## Quieting the messages

Every function accepts `silent = TRUE`, and
`options(crosswalk.silent = TRUE)` silences the whole package. Silencing
suppresses messages only -- the `crosswalk_metadata` and `join_quality`
attributes are always attached.

## Learn more

- `vignette("how-interpolation-works")` -- how allocation factors are applied,
  choosing a `weight`, the approximation involved in non-count variables, and
  diagnosing join quality
- `vignette("county-crosswalks")` -- county redefinitions and arbitrary-year
  county crosswalks
- `vignette("standardizing-longitudinal-data")` -- a worked example building a
  six-year tract panel from mixed-vintage data
