---
title: "How Interpolation Works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How Interpolation Works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
## the accuracy-assessment section pulls data from the Census API
has_census_key <- Sys.getenv("CENSUS_API_KEY") != ""

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5)
```

Crosswalking produces *estimates*: when a source geography is split across
several target geographies, its values must be apportioned, and every
apportionment rests on assumptions. This vignette explains exactly what
`crosswalk_data()` computes, how to choose the `weight` that drives it, and
how to assess the quality of the results.

```{r setup}
library(crosswalk)
library(dplyr)
library(tibble)
```

## Allocation factors

Every crosswalk relates source geographies to target geographies with an
*allocation factor*: the share of the source unit's weighting attribute
(population, housing units, or land area) that falls inside each target unit.
Consider a miniature crosswalk in which source unit "A" lies entirely inside
target "X", while source unit "B" is split between "X" (a quarter of B's
population) and "Y" (three quarters):

```{r toy-crosswalk}
toy_crosswalk <- tribble(
  ~ source_geoid, ~ target_geoid, ~ allocation_factor_source_to_target,
  "A",            "X",            1,
  "B",            "X",            0.25,
  "B",            "Y",            0.75)
```

Allocation factors for each source unit sum to 1 (all of a source unit's
weight lands somewhere), and the crosswalk has exactly one row per
source-target pair.

## Interpolating counts

For count variables (population, households, loan originations, ...), each
source value is multiplied by the allocation factor and the products are
summed within each target geography:

$$\text{count}_{\text{target}} = \sum_{\text{sources}} \text{count}_{\text{source}} \times \text{allocation factor}_{\text{source} \to \text{target}}$$

```{r toy-counts}
toy_data <- tribble(
  ~ source_geoid, ~ count_people,
  "A",            100,
  "B",            400)

crosswalk_data(
  data = toy_data,
  crosswalk = toy_crosswalk,
  silent = TRUE)
```

Target X receives all of A plus a quarter of B (100 + 0.25 x 400 = 200), and
target Y receives the remaining three quarters of B (300). The total (500) is
preserved: counts are neither created nor destroyed, only reapportioned.

The embedded assumption: the variable being crosswalked is distributed across
the source unit *proportionally to the weighting attribute*. Allocating a
quarter of B's loan originations to X is exactly right only if B's loans are
distributed like B's population. This is why matching the `weight` to your
variable matters (see below) -- and why results for small areas should be
treated as estimates.

## Interpolating non-count variables

Means, medians, percentages, and rates can't be summed. For these,
`crosswalk_data()` computes a *weighted mean* of the source values within each
target geography, using the allocation factors as weights:

```{r toy-noncounts}
toy_medians <- tribble(
  ~ source_geoid, ~ median_income,
  "A",            50000,
  "B",            100000)

crosswalk_data(
  data = toy_medians,
  crosswalk = toy_crosswalk,
  silent = TRUE)
```

Target X's value is (1 x 50000 + 0.25 x 100000) / 1.25 = 60000.

Two approximations are involved:

1. **A weighted mean of medians is not a median.** The true median income of
   X's residents depends on the full income distributions of A and B, which
   the crosswalk does not carry. The same applies to percentages: the exact
   percentage for X would weight each source's percentage by its
   *denominator* population in X, which is only approximated here.
2. **Allocation factors are shares, not sizes.** The factor 0.25 says "a
   quarter of B goes to X" -- it does not say how large B is relative to A. If
   B has ten times A's population, B's quarter inside X contains far more
   people than all of A, but the weighted mean gives A four times B's weight
   (1 vs. 0.25). The approximation is most accurate when source units are of
   broadly similar size, as census units within a geography type generally
   are.

When precision matters for a non-count variable, a more accurate route is to
crosswalk its *components* as counts (numerator and denominator separately)
and recompute the ratio at the target level.

## Choosing a weight

The `weight` argument controls which attribute the allocation factors are
built from:

| `weight` | Allocates by | Use for |
|---|---|---|
| `"population"` (default) | Resident population in the overlap | People-denominated variables: population counts, poverty, demographic shares |
| `"housing"` | Housing units in the overlap | Housing-stock variables: units, tenure, vacancy, home loans |
| `"land"` | Land area of the overlap | Land-based variables, or when settlement patterns are irrelevant |

Guidance:

- Match the weight to the variable's denominator. Housing-unit weights
  allocate a tract's housing stock more faithfully than population weights do
  (group quarters hold people but no housing units, for example).
- Population and housing weights are usually similar; land-area weights can
  differ dramatically in sparsely settled areas (a target unit covering 90% of
  a source's land might hold 5% of its people).
- **NHGIS caveat**: block-based NHGIS crosswalks publish a single combined
  interpolation weight, so `weight` does not vary the allocation factors for
  those crosswalks. Tract- and block-group-level NHGIS crosswalks do provide
  distinct population and housing weights. The `weighting_factor` column of
  the returned crosswalk always records what was actually used.
- County-events crosswalks use `"identity"` weights for unchanged and renamed
  counties (factor 1) and population-based factors for splits and part
  transfers, regardless of `weight`.

## Assessing accuracy against published values

For geographies the Census Bureau publishes directly, we can check
crosswalked estimates against observed values. Here we crosswalk ZCTA-level
poverty counts to PUMAs, then compare against the ACS's own PUMA-level
figures. (This section queries the Census API, so it requires a
`CENSUS_API_KEY`.)

```{r accuracy-setup, eval = has_census_key}
zcta_poverty <- tidycensus::get_acs(
    year = 2023,
    geography = "zcta",
    output = "wide",
    variables = c(below_poverty = "B17001_002"),
    progress_bar = FALSE) |>
  select(
    source_geoid = GEOID,
    count_below_poverty = below_povertyE)

puma_poverty_crosswalked <- crosswalk_data(
  data = zcta_poverty,
  source_geography = "zcta",
  target_geography = "puma",
  weight = "population",
  silent = TRUE)
```

```{r accuracy-compare, eval = has_census_key}
library(ggplot2)

puma_poverty_published <- tidycensus::get_acs(
    year = 2023,
    geography = "puma",
    output = "wide",
    variables = c(below_poverty = "B17001_002"),
    progress_bar = FALSE) |>
  select(
    puma_geoid = GEOID,
    count_below_poverty_published = below_povertyE)

comparison <- puma_poverty_published |>
  left_join(puma_poverty_crosswalked, by = c("puma_geoid" = "geoid")) |>
  mutate(
    difference_percent =
      (count_below_poverty_published - count_below_poverty) /
      count_below_poverty_published)

comparison |>
  ggplot() +
    geom_histogram(aes(x = difference_percent), bins = 60) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    scale_x_continuous(labels = scales::percent) +
    labs(
      title = "Crosswalked data approximates observed values",
      subtitle = "ZCTA-level poverty counts crosswalked to PUMAs vs. published PUMA values",
      y = "Number of PUMAs",
      x = "Percent difference between published and crosswalked values")
```

Most PUMAs land within a few percent of their published values. The residual
error comes from the proportional-distribution assumption above -- and it
shrinks as source units get smaller relative to target units. Crosswalking
from block groups instead of ZCTAs would track published values more closely;
crosswalking from large units to small ones (counties to tracts, say) deserves
much more skepticism than the reverse.

Also note both sides of this comparison are survey estimates with their own
margins of error. The package does not propagate ACS margins of error through
interpolation.

## Diagnosing join quality

Interpolation can only be as good as the join between your data and the
crosswalk. Data GEOIDs missing from the crosswalk are *dropped* (their values
allocate nowhere), so `crosswalk_data()` reports on the join as it runs and
attaches the full statistics to its result as the `join_quality` attribute --
computed even when messages are silenced:

```{r join-quality, eval = has_census_key}
join_quality <- attr(puma_poverty_crosswalked, "join_quality")

## share of data GEOIDs that failed to match the crosswalk (dropped)
join_quality$pct_data_unmatched

## the unmatched GEOIDs themselves
head(join_quality$data_geoids_unmatched)

## share of crosswalk GEOIDs with no data (usually benign: geographies
## with no observations in your data)
join_quality$pct_crosswalk_unmatched
```

A small unmatched share is common -- GEOID vintages differ subtly across
products, and some datasets use placeholder codes. A *large* share usually
means a vintage mismatch: e.g., 2023-vintage tract data joined to a crosswalk
built for 2010s-era tracts. Checking a few unmatched GEOIDs against the
[Census geocoder](https://geocoding.geo.census.gov/) or the source data's
documentation usually identifies the problem quickly.

For state-nested geographies (counties, tracts, block groups, blocks), the
attribute also reports whether unmatched GEOIDs concentrate in particular
states (`state_analysis_data`; see `?crosswalk_data`). Concentration is a
diagnostic clue: unmatched GEOIDs spread thinly across many states suggest
scattered data quirks, while a pile-up in one state suggests a systematic
boundary or vintage issue (Connecticut's 2022 county change is a frequent
culprit).

ZCTAs cross state lines, so state analysis doesn't apply to the example
above. For non-nested geographies, joining unmatched GEOIDs to spatial data is
an effective way to look for patterns:

```{r spatial-diagnosis, eval = FALSE, purl = FALSE}
## e.g., where are unmatched ZCTAs concentrated? (requires sf + tigris)
zctas_sf <- tigris::zctas(year = 2023, progress_bar = FALSE)
states_sf <- tigris::states(year = 2023, cb = TRUE, progress_bar = FALSE)

zctas_sf |>
  dplyr::filter(GEOID20 %in% join_quality$data_geoids_unmatched) |>
  sf::st_intersection(states_sf |> dplyr::select(NAME)) |>
  sf::st_drop_geometry() |>
  dplyr::count(NAME, sort = TRUE)
```

When we ran this diagnosis on the ZCTA -> PUMA example, unmatched ZCTAs were
roughly proportional to state populations -- except for a disproportionate
cluster in Washington, DC corresponding to federal areas and buildings: real
ZCTAs, but ones with no place in a population-weighted crosswalk.

## Summing up

- Counts are apportioned by allocation factor and summed; totals are
  preserved.
- Non-counts are approximated by allocation-factor-weighted means; when
  precision matters, crosswalk numerators and denominators as counts instead.
- Match `weight` to your variable's denominator.
- Check the `join_quality` attribute -- unmatched data rows are dropped, and a
  concentrated pattern of unmatched GEOIDs usually signals a vintage mismatch.
- Prefer small source units relative to target units; treat the reverse with
  skepticism.
